Formalized Linear Programming

### 2019-10-06 TeX/paper.tex
Fix empty lines after notes
We lack related work
update paper

\section{Linear Programming}
\color{red} The standard simplex algorithm. No reference to Lean code \color{black}

The simplex algorithm aims to optimize the value of an objective function within the constraints of a set of affine inequalities. It can be used to solve problems of the following form.
% \color{red} This is a slightly nonstandard way of writing a linear program. The usual way would be $Ax \le b$. However the matrix that is stored during a simplex is $-A$ according to that representation, but $A$ according to the representation below.\color{black}
\begin{equation}
  \begin{aligned}
    \text{Find } x \text{ that maximizes } && c^Tx \\
    \text{subject to } && A x + b \ge 0\\
    \text{and } && x \ge 0
  \end{aligned}
\end{equation}
Here, $x \in \mathbb{Q}^n$, $c \in \mathbb{Q}^n$, $b \in \mathbb{Q}^m$, and $A \in \mathbb{Q}^{m \times n}$.
This can be written in an equivalent form \color{red} The inclusion of $z$ in the objective function is slightly unusual because obviously maximizing $c^Tx$ is the same as maximizing $c^Tx + z$. However $z$ is a cell in the tableau, it is the objective function value of the sample solution, and the explanation of the tableau is slightly easier with it\color{black}
\begin{equation}
  \begin{aligned}
    \text{Find } x_r \text{ and } x_c \text{ that maximize } && c^Tx_c + z \\
    \text{subject to } && x_r = A x_c + b \\
    \text{and } && x_r \ge 0, x_c \ge 0
  \end{aligned}
\end{equation}
with $x_r \in \mathbb{Q}^m$, $x_c \in \mathbb{Q}^n$ and $z \in \mathbb{Q}$

@@

In the tableau representation of this problem, the vectors $x_r$ and $x_c$ are viewed as minors of a single variable $x$. The subset of $\mathbb{Q}^{m+n}$, $\{x : x_c = Ax_r + b \wedge x_c \ge 0 \wedge x_r \ge 0\}$ is stored as part of the simplex algorithm. The tableau consists of a matrix $A$ of coefficients, a column vector $b$, a reduced cost vector $c$, and the constant part of the objective function $z$. Additionally two vectors of indices $r$ and $c$ defining the minors $x_r$ and $x_c$ are also stored.
\begin{equation}
  \begin{aligned}
    x_{r_1} && = && A_{11}x_{c_1} && + && A_{12}x_{c_2} && + && \dots && + && A_{1n}x_{c_n} && + && b_1 \\
    x_{r_2} && = && A_{21}x_{c_1} && + && A_{22}x_{c_2} && + && \dots && + && A_{2n}x_{c_n} && + && b_2 \\
    \vdots && && \vdots && && \vdots && && && && \vdots && \vdots \\
    x_{r_m} && = && A_{m1}x_{c_1} && + && A_{m2}x_{c_2} && + && \dots && + && A_{mn}x_{c_n} && + && b_n \\
    c^Tx_c + z && = && c_1x_{c_1} && + && c_2x_{c_2} && + && \dots && + && c_nx_{c_n} && + && z
  \end{aligned}
\end{equation}
In this tableau representation the $x_r$ are known as the \textit{row} variables, and the $x_c$ are the \textit{column} variables. In a tableau the row variables are expressed as affine functions in terms of the column variables.

@@

The choice of pivot row and column satisfying (\ref{eqn:valid_pivot}) is non-unique, and must be taken carefully in order to guarantee termination. Different implementations of the simplex algorithm use different pivot rules to guarantee termination. If $z$ was strictly increasing it would be straightforward to prove the algorithm terminates, since there are only finitely many partitions of the variables into row and column variables, and the simplex algorithm cannot repeat a partition if $z$ is strictly increasing. However when $b_i = 0$ in the pivot row $i$, $z$ does not increase, and ensuring termination is more difficult. Our implementation uses Bland's Rule~\cite{Blands_rule} to ensure termination.

@@

The Lean simplex algorithm takes as arguments a tableau, the row that should be optimized and a boolean function, that gives the user the option of terminating the algorithm early. This can be used for example to determine whether the maximum value of a variable is positive, without actually computing the maximum. The simplex algorithm itself returns both a tableau and the reason for termination of the algorithm, either \lstinline|unbounded|, \lstinline|optimal|, or \lstinline|while|. The
\begin{lstlisting}
def simplex (w : tableau m n → bool) (obj : fin m) : Π (T : tableau m n) (hT : feasible T),
  tableau m n × termination n
\end{lstlisting}

\lstinline|termination n| is an inductive type with three constructors. In the case that the variable is unbounded, the piivot column that was chosen before the algorithm failed to find a pivot row is also returned.
\begin{lstlisting}
inductive termination (n : ℕ) : Type
| while {}              : termination
| unbounded (c : fin n) : termination
| optimal {}            : termination
\end{lstlisting}

@@

Various properties of the tableau then have to be proven by induction on this relation. The majority of the length of the termination proof was taken up by proving properties of related tableaux and tableaux within a cycle.

@@

The simplex algorithm was originally written on an extremely slow implementation of matrices. The implementation in the Lean maths library \color{red}[cite mathlib]\color{black} is as the type of binary functions from two finite sets into a ring, in our case $\mathbb{Q}$. This implementation is convenient for writing proofs, but very slow to compute. In order to have a practical executable version of the simplex algorithm, it was necessary to transfer the algorithm and proofs onto a faster implementation of matrices using arrays. Ideally this would be done using a refinement framework \color{red} brief explanation of what that means here. Would it be possible to generalize this proof onto a type class rather than a concrete type using this framework?\color{black}. This is not currently available in Lean, so it was instead done in an ad hoc way, using a tableau type class. The proof of the simplex algorithm can be generalized onto any type for which there is an instance of the \lstinline|is_tableau| type class.

>>>>

\section{Linear Programming}
\color{red} The standard simplex algorithm. No reference to Lean code \color{black}
The simplex algorithm aims to optimize the value of an objective function within the constraints of a set of affine inequalities. It can be used to solve problems of the following form.
% \color{red} This is a slightly nonstandard way of writing a linear program. The usual way would be $Ax \le b$. However the matrix that is stored during a simplex is $-A$ according to that representation, but $A$ according to the representation below.\color{black}
\begin{equation}
  \begin{aligned}
    \text{Find } x \in \mathbb{Q}^n \text{ that maximizes } && c^Tx \\
    \text{subject to } && A x + b \ge 0\\
    \text{and } && x \ge 0
  \end{aligned}
\end{equation}
Here, $x \in \mathbb{Q}^n$, $c \in \mathbb{Q}^n$, $b \in \mathbb{Q}^m$, and $A \in \mathbb{Q}^{m \times n}$.
This can be written in an equivalent form \color{red} The inclusion of $z$ in the objective function is slightly unusual because obviously maximizing $c^Tx$ is the same as maximizing $c^Tx + z$. However $z$ is a cell in the tableau, it is the objective function value of the sample solution, and the explanation of the tableau is slightly easier with it\color{black}
\begin{equation}
  \begin{aligned}
    \text{Find } x_r \text{ and } x_c \text{ that maximize } && c^Tx_c + z \\
    \text{subject to } && x_r = A x_c + b \\
    \text{and } && x_r \ge 0, x_c \ge 0
  \end{aligned}
\end{equation}
with $x_r \in \mathbb{Q}^m$, $x_c \in \mathbb{Q}^n$ and $z \in \mathbb{Q}$

@@

In the tableau representation of this problem, the vectors $x_r$ and $x_c$ are viewed as minors of a single variables $x$. The subset of $\mathbb{Q}^{m+n}$, $\{x : x_c = Ax_r + b \wedge x_c \ge 0 \wedge x_r \ge 0\}$ is stored as part of the simplex algorithm. The tableau consists of a matrix $A$ of coefficients, a column vector $b$, a reduced cost vector $c$, and the constant part of the objective function $z$. Additionally two vectors of indices $r$ and $c$ defining the minors $x_r$ and $x_c$ are also stored.
\begin{equation}
  \begin{aligned}
    x_{r_1} && = && A_{11}x_{c_1} && + && A_{12}x_{c_2} && + && \dots && + && A_{1n}x_{c_n} && + && b_1 \\
    x_{r_2} && = && A_{21}x_{c_1} && + && A_{22}x_{c_2} && + && \dots && + && A_{2n}x_{c_n} && + && b_2 \\
    \vdots && && \vdots && && \vdots && && && && \vdots && \vdots \\
    x_{r_m} && = && A_{m1}x_{c_1} && + && A_{m2}x_{c_2} && + && \dots && + && A_{mn}x_{c_n} && + && b_n \\
    c^Tx_c + z && = && c_1x_{c_1} && + && c_2x_{c_2} && + && \dots && + && c_nx_{c_n} && + && z
  \end{aligned}
\end{equation}
In this tableau representation the $x_r$ are known as the \textit{row} variables, and the $x_c$ are the \textit{column} variables. In a tableau the row variables are expressed as affine functions in terms of the column variables.

@@

The choice of pivot row and column satisfying (\ref{eqn:valid_pivot}) is non-unique, and must be chosen carefully in order to guarantee termination. Different implementations of the simplex algorithm use different pivot rules to guarantee termination. If $z$ was strictly increasing it would straightforward to prove the algorithm terminates, since there are only finitely many partitions of the variables into row and column variables, and the simplex algorithm cannot repeat a partition if $z$ is strictly increasing. However when $b_i = 0$ in the pivot row $i$, $z$ does not increase, and ensuring termination is more difficult. Our implementation uses Bland's Rule to ensure termination. \cite{Blands_rule}.

@@

The Lean simplex algorithm takes as arguments a tableau, the row that should be optimized and a boolean function, that gives the user the option of terminating the algorithm early. This can be used for example to determine whether the maximum value of a variable is positive, without actually computing the maximum. The simplex algorithm itself returns both a tableau and the reason for termination of the algorithm, either \lstinline|unbounded|, \lstinline|optimal|, or \lstinline|while|. The function takes as an argument a proof of feasibility of the tableau, since it does not terminate or return a useful for non feasible tableau.
\begin{lstlisting}
def simplex (w : tableau m n → bool) (obj : fin m) : Π (T : tableau m n) (hT : feasible T),
  tableau m n × termination n
\end{lstlisting}

\lstinline|termination n| is an inductive type with three constructors. In the case that the variable is unbounded, the pivot column that was chosen before the algorithm failed to find a pivot row is also returned.
\begin{lstlisting}
inductive termination (n : ℕ) : Type
| while {}              : termination
| unbounded (c : fin n) : termination
| optimal {}            : termination
\end{lstlisting}

@@

Various properties of the tableau then have to be proven by induction on this relation. The majority of the length of the termination proof was taken up by proving properties of related tableaux and tableaux within a cycle. For example we proved that if a column is owned by a different variable in two related tableau \lstinline|T| and \lstinline|T'|, then this column was the pivot column at some tableau visited in between \lstinline|T| and \lstinline|T'|. Also proven was the fact that if the objective function value at the sample point does not change between \lstinline|T| and \lstinline|T'|, then neither does any value in the constant column. These were then used to prove the relation was irreflexive, and therefore that the algorithm does not repeat a tableau.

@@

The simplex algorithm was originally written on an extremely slow implementation of matrices. The implementation in the Lean maths library \color{red}[cite mathlib]\color{black} is as the type of binary functions from two finite sets into a ring, in our case $\mathbb{Q}$. This implementation is convenient for writing proofs, but very slow to compute. In order to have a practical executable version of the simplex algorithm, it was necessary to transfer the algorithm and proofs onto a faster implementation of matrices using arrays. Ideally this would be done using a refinement framework \color{red} brief explanation of what that means here. Would it be possible to generalize proof onto a type class rather than a concrete type using this framework?\color{black}. This is not currently available in Lean, so it was instead done in an ad hoc way, using a tableau type class. The proof of the simplex algorithm can be generalized onto any type for which there is an instance of the \lstinline|is_tableau| type class.
